Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-01-19 paper/paper.Rmd
Some rewording.

\textcolor{blue}{Here we are providing a solution to fix a problem that has been forever present in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates the crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of points are sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equally represented in the low-dimensional space. It places focus on the space, that volume is spread evenly, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}

@@

<!--In this work we address data crowding in low-dimensional linear projections by providing a reverse transformation for tour methods. Tours show interpolated sequences of low-dimensional projections of the data. When exploring data with a tour we can discover features that are only visible in linear combinations of variables. However, the data crowding could hide these features. To reverse the effect, we introduce a radial transformation that magnifies the center of the distribution. This is called a burning sage tour, to reflect that the crowding caused by the curse of dimensionality is being removed.-->

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, not dimension reduction, problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches would still be applied though after the transformation, especially for large sample sizes.}

@@

```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom). Three selected clusters are highlighted in color, all other points are shown in grey. Using the sage display mitigates overplotting and provides a better understanding of cluster separation.", out.width="0.9\\textwidth", fig.align="center"}
library(magick)
dir <- here::here("../pngs/") # dir <- here::here("pngs//")
frames <- c("001", "016", "038", "074")
draw_frames <- function(pngs) {
  res <- lapply(pngs, image_read)
  res <- do.call("c", res)
  image_append(res)
}

>>>>

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of points are sampled uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$) are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, not dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches can be still be applied after the transformation, especially for large sample sizes.}

@@

```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom). Three selected clusters are highlighted in color, all other points are shown in grey. Using the sage display mitigates overplotting and provides a better understanding of cluster separation.", out.width="0.9\\textwidth", fig.align="center"}
library(magick)
dir <- here::here("../pngs/")
#dir <- here::here("pngs//")
frames <- c("001", "016", "038", "074")
draw_frames <- function(pngs) {
  res <- lapply(pngs, image_read)
  res <- do.call("c", res)
  image_append(res)
}

### 2021-01-19 paper/paper.tex
Some rewording.

\textcolor{blue}{Here we are providing a solution to fix a problem that has been forever present in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates the crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
points are sampled uniformly within \(p\)-dimensional hyperspheres
(\(p=3, 10, 100\)) are displayed as hexbin plots. Color indicates log
count of the bin, with yellow mapping the highest counts. As \(p\)
increases the density concentrates in the center of the projection.
\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equally represented in the low-dimensional space. It places focus on the space, that volume is spread evenly, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, not dimension reduction, problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches would still be applied though after the transformation, especially for large sample sizes.}

>>>>

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
points are sampled uniformly within \(p\)-dimensional hyperspheres
(\(p=3, 10, 100\)) are displayed as hexbin plots. Color indicates log
count of the bin, with yellow mapping the highest counts. As \(p\)
increases the density concentrates in the center of the projection.
\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the right-most plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays [@chang]. These are designed to tackle large sample size, not dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches can be still be applied after the transformation, especially for large sample sizes.}
